---
title: "How many particles must I measure for a desired precision?"
author: "John Minter"
date: "2018-03-13"
bibliography: ./inc/psd.bib
csl: ./inc/acs.csl
output:
  pdf_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

A typical image analysis measurement for a particle size distribution on
existing systems measures from 1000 to 2500 individual particles. Many
scientists have asked how many particles need to be measured to achieve
a particular precision. We demonstrate in this section that, for
broad distributions, significantly larger particle counts are required.

# Masuda and Iinoya's model

Masuda and Iinoya (@Masuda1971a) developed a theoretical model to answer
this question. The model assumes the particle size distribution is
lognormal and that there is negligible interaction between particles.

Masuda and Iinoya assume that the observable quantity, $y$, is related
to the particle diameter, $D_p$, by a power law relation:

$$y = K D_{p}^{\alpha}$$

where $K$ is a constant. The basis, $\beta$, of the distribution is 0
for a count-basis and 3 for a mass basis. Masuda and Iinoya define a
parameter, $c$, that is usually denoted by $J$ at Kodak, such that

$$J = c \equiv \beta + \frac{\alpha}{2}$$

Masuda and Iinoya calculate the probability, $P$, that the average
diameter has a mean error, $\epsilon$, less than some specified relative
error, $\delta$. These authors note that

$$\Phi(-|u|) = \frac{1-P}{2}$$

where

$$ \Phi(u) \equiv \frac {1} {\sqrt {2\pi}} \int_{-\infty}^{u} e^{-\frac{u^{2}}{2}}du $$

Masuda and Iinoya provide the following table for $u$ as a function of
the desired probability, $P$.

\begin{table}[]
\centering
\caption{The parameter, $u$, as a function of the desired probability, $P$. Data from Masuda and Iinoya.}
\label{my-label}
\begin{tabular}{lllllllllll}
\textbf{P} & 0.50 & 0.75 & 0.80 & 0.90 & 0.95 & 0.975 & 0.990 & 0.995 & 0.998 & 0.999 \\
\textbf{u} & 0.67 & 1.15 & 1.28 & 1.64 & 1.96 & 2.24  & 2.58  & 2.81  & 3.09  & 3.29
\end{tabular}
\end{table}

Masuda and Iinoya compute the parameter $\sigma$ from the
geometric standard deviation of the distribution


$$ \sigma = \ln(GSD) $$

and compute the parameter, $\omega$

$$ \omega = u^{2} \alpha^{2} \sigma^{ 2} \left(2c^{2} \sigma^{ 2} + 1\right) $$

that finally permits calculation that the number of particles, $n^{*}$,
required to measure the mean particle diameter to a given precision, as
shown below:

$$ \log \left(n^{*} \right) = -2\log(\delta) + \log(\omega) $$

This calculation is easily performed using an Excel97 worksheet for a
desired precision (specified by $P$ and $\delta$) and a given moment of
the distribution (specified by choices of $\alpha, \beta$, and,
therefore, $c$ or $J$) and a specified geometric standard deviation.

The essential result of this theoretical treatment is that $n^*$
increases rapidly with increasing geometric standard deviation
(i.e. width) of the distribution and the particular moment of the
distribution that is desired (given by $c$ or $J$).

We suggest that the `natural moment' of EIA is J = 0.5, although
typically J = 0 is computed from the lognormal fit using a histogram
with bins equally spaced in $ln(diameter)$. Note that analysis by
Summa averages projected area (instead of equivalent circular diameter)
and has a ``natural moment'' of J = 1. Note that significantly larger
counts are required to measure the J = 3 moment with equivalent
precision.

Because the distribution width of Kodak emulsions is variable and the
distribution parameters required to answer a particular client’s
question vary, these results suggest that the number of
particles analyzed needs to be determined on a case-by-case basis.
The Excel97 spreadsheet developed makes such calculations easy.


I am now (2018-03-14) in the process of converting this from Excel to R.
The functions to convert the confidence levels to the required Z
values were added to the **rAnaLab** package.

```{r, computeUandP}
suppressPackageStartupMessages(library(rAnaLab))

conf.levs <- c(0.500, 0.750, 0.800, 0.900, 0.950, 
               0.975, 0.990, 0.995, 0.998, 0.999)

z.vals <- unlist(lapply(conf.levs, cl.to.z))
print(z.vals)
```


We will now create a datafame using Masuda and Iinoya's nomenclature
and plot the relationship between $P$ and $u$,

```{r, message=FALSE}
suppressPackageStartupMessages(library(ggplot2))

df <- data.frame(P=conf.levs, u=z.vals)
lSize <- 1
plt <- ggplot() +
       geom_point(data=df, aes(x=P, y=u),  colour="darkblue") + 
       geom_line(data=df, aes(x=P, y=u), colour='darkblue',
                 size=lSize) +
       xlab("P") +
       ylab("u") +
       ggtitle("Masuda and Iinoya's u(P)") +
       theme(axis.text=element_text(size=12),
             axis.title=element_text(size=12),
            plot.title=element_text(hjust = 0.5)) # center the title
print(plt)

```

We will assume $u = 0.05$ (5%). We will compute for
c (or J) = 0.5, 1, and 3. Because we are doing image analysis and
counting individual particles, $\beta = 0$ so $\alpha = 2J$.

So, we will create a vector of geometric standard deviation values
and use this to create a dataframe.

```{r, createDataFrame}
gsd <- seq(from=1.01, to=2.49, by=0.01)
sigma <- log(gsd)
u <- 1.9599640
delta <- 0.01
J <- 0.5
alph <- 2.0*J
omega = u^2*alph^2*sigma^2*(2*J^2*sigma^2+1)
lgNstar <- - 2.0*log(delta)+log(omega)
nStarP5 <- exp(lgNstar)
J <- 1.0
alph <- 2.0*J
omega = u^2*alph^2*sigma^2*(2*J^2*sigma^2+1)
lgNstar <- - 2.0*log(delta)+log(omega)
nStar1 <- exp(lgNstar)

J <- 3.0
alph <- 2.0*J
omega = u^2*alph^2*sigma^2*(2*J^2*sigma^2+1)
lgNstar <- - 2.0*log(delta)+log(omega)
nStar3 <- exp(lgNstar)

df <- data.frame(gsd=gsd, nStarP5=nStarP5,
                 nStar1=nStar1, nStar3=nStar3)

print(summary(df))
```

A plot of the number of particles required to measure selected moments
of the distribution is shown below.

```{r, plotResults, fig.cap="The number of particles required to measure the j^th^ moment of the diameter distribution, D~J~, to a precision of 0.5% at the 0.995 confidence interval for selected values of J. Values calculated using the model of Masuda and Iinoya."}

suppressPackageStartupMessages(library(ggplot2))

plt <- ggplot(data=df) +
       geom_line(aes(x=gsd, y=nStarP5), colour="red") +
       geom_line(aes(x=gsd, y=nStar1), colour="blue") +
       geom_line(aes(x=gsd, y=nStar3), colour="darkblue") +
       scale_x_continuous(breaks = seq(from=1.0, to=2.5, by=0.1),
                          limits = c(1.0,2.5)) +
       scale_y_log10(breaks = c(10, 10^2, 10^3, 10^4, 10^5,
                                10^6, 10^7, 10^8),
                     limits = c(1.0,1e08)) +
       xlab(label="geometric standard deviation") +
       ylab(label="# particles") +
       ggtitle("") +
       theme(axis.text=element_text(size=12),
             axis.title=element_text(size=12),
             # center the title
             plot.title = element_text(hjust = 0.5))

print(plt)

```

These values agree quite well with those computed using Roger Button's
spreadsheet in 1999. That workbook is no longer available but the plot
is included in the technical report we wrote at the time. One advantage
of performing the computation in R is that functions that may be reused
are easily included in packages under version control and the Rmarkdown
documents make it easy to "show our work."
