---
title: "Models for mass thickness contrast in transmission electron imaging"
author: "John Minter"
date: "2018-03-13"
bibliography: ./inc/psd.bib
csl: ./inc/acs.csl
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The theory of mass thickness contrast in transmission electron imaging
was developed by Smith and Burge [@Smith1963a] refined by Misell and
Burdett [Misell1977a]. The most important aspect of this theory is that
the plural scattering cross section for electrons, $S_{P}$, is
reasonably insensitive to atomic number but is dependent upon the
accelerating voltage, $V_{0}$, and the aperture half angle, $\alpha$.
The ratio of the image intensity through the specimen, $I$, to the
incident electron intensity, $I_{0}$, is given by:

$$ \frac{I}{I_{0}} = e^{-S_{P}\left(V_{0},\alpha \right) \cdot p \cdot t } $$
where $\rho$ is the density and $t$ is the thickness. If $rho$ has units
of $g/cm^3$ and $t$ has units of $cm$, then $S_{P}$ has units of
$cm^{2}/g$. Misell and Burdett provided tabulated values for selected
values of $S_{P}(V0,\alpha)$. Note that Misell and Burdett reported the
data in units of $m^2/mg$, so we will need to transform the data.

These values do not cover all the voltages and aperture half angles of
interest. We noticed that a plot of ln(SP) vs. ln(V0) is linear, as
shown below.

First, we load the raw data, taken from p. 175 of the paper.

```{r, loadDF, message=FALSE}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(pander))
suppressPackageStartupMessages(library(tidyverse))
fi <- here('dat/Misell-Burdett.csv')
df <- read_csv(fi)
pander(df)
```

\clearpage

Next, transform the data by:

1. Convert $kV$ to $eV$ (fac1)
2. Convert $m^2/mg$ to $cm^2/g$ (fac2)
3. Take the natural log of each term. *N.B.* In R the natural log
function is `log()` while the base 10 function is `log10()`.


```{r, makeLogDF}
fac1 <- 1000 # kev to eV
fac2 <- 10000000 # m2/mg to cm2/g 
df.ln <- data.frame(ln.E0.eV=log(fac1*df[,1]),
                    ln.alpha.05=log(fac2*df[,2]),
                    ln.alpha.10=log(fac2*df[,3]),
                    ln.alpha.15=log(fac2*df[,4]))
names(df.ln) <- c("ln.e0.eV","ln.alpha.05",
                  "ln.alpha.10","ln.alpha.15")
pander(df.ln)
```

Next, let'd consider the summary statistics of the data.

```{r, summarizeLnDf}
pander(summary(df.ln))
```

Next, let's fit the data to a linear regression for each aperture,
starting with the 5 mrad aperture.

```{r, fitAp05}
x.pred <- seq(from=9.90, to=11.6, by=0.01)

df.05 <- data.frame(x=df.ln[,1], y=df.ln[,2])
lm.05 <- lm(y~x, data=df.05)
sum.05 <- summary(lm.05)
coe.05 <- sum.05$coef
pander(coe.05)
```

\clearpage 

We now define a function to get the slope and intercept from the
coefficients so we can tabulate them.

```{r, getCoef05}
getInterceptFromCoef <- function(coe){
  mu <- coe[1]
  se <- coe[3]
  ret <- c(mu,se)
  names(ret) <- c("mu", "se")
  return(ret)
}

getSlopeFromCoef <- function(coe){
  mu <- coe[2]
  se <- coe[4]
  ret <- c(mu,se)
  names(ret) <- c("mu", "se")
  return(ret)
}

sl.05 <- getInterceptFromCoef(coe.05)
in.05 <- getSlopeFromCoef(coe.05)
```
And compute the predicted values for the line

```{r, predAp05}
df.05.p <- data.frame(x=x.pred)
df.05.p$y <- predict(lm.05, newdat=df.05.p)
```

We repeat the process for the 10 mrad aperture.

```{r, fitAp10}
df.10 <- data.frame(x=df.ln[,1], y=df.ln[,3])
lm.10 <- lm(y~x, data=df.10)
sum.10 <- summary(lm.10)
coe.10<- sum.10$coef
sl.10 <- getInterceptFromCoef(coe.10)
in.10 <- getSlopeFromCoef(coe.10)
pander(coe.10)
```

Again, computing the predicted values.

```{r, predAp10}
df.10.p <- data.frame(x=x.pred)
df.10.p$y <- predict(lm.10, newdat=df.10.p)
```

Finally, we fit the data for the 15 mrad aperture.

```{r, fitAp15}
df.15 <- data.frame(x=df.ln[,1], y=df.ln[,4])
lm.15 <- lm(y~x, data=df.15)
sum.15 <- summary(lm.15)
coe.15<- sum.15$coef
coe.10<- sum.10$coef
sl.15 <- getInterceptFromCoef(coe.15)
in.15 <- getSlopeFromCoef(coe.15)
pander(coe.15)
```

computing the predicted values as before.

```{r, predAp15}
df.15.p <- data.frame(x=x.pred)
df.15.p$y <- predict(lm.15, newdat=df.15.p)
```

\clearpage

Next, we plot the data.

```{r, makeLogPlot}

plt <- ggplot() +
       geom_point(data=df.ln, aes(x=ln.e0.eV, y=ln.alpha.05),
                  colour="darkblue") +
       geom_line(data=df.05.p, aes(x=x, y=y), colour="darkblue")  +
       geom_point(data=df.ln, aes(x=ln.e0.eV, y=ln.alpha.10),
                  colour="blue") +
       geom_line(data=df.10.p, aes(x=x, y=y), colour="blue")  +
       geom_point(data=df.ln, aes(x=ln.e0.eV, y=ln.alpha.15),
                  colour="red") +
       geom_line(data=df.15.p, aes(x=x, y=y), colour="red")  +
       xlab(label="ln(e0 [eV])") +
       ylab(label="ln(Sp [cm2/g])") +
       ggtitle("Misell-Burdett data") +
       theme(axis.text=element_text(size=12),
             axis.title=element_text(size=12),
             # center the title
             plot.title = element_text(hjust = 0.5))

print(plt)
```

\clearpage

Now we can aggregate the data

```{r, aggregateCoef}
slope <- round(rbind(sl.05, sl.10, sl.15), 4)
inter <- round(rbind(in.05, in.10, in.15), 4)
mrad <- c(5,10,15)
mat.fit <- cbind(mrad, slope, inter)

df.fit <- as.data.frame(matrix(unlist(mat.fit), nrow = 3))
names(df.fit) <- c("mrad", "slope.mu", "slope.se",
                   "inter.mu", "inter.se")
pander(df.fit)
```

\clearpage

# References
